---
title: "survival_prediction"
output:
  html_document: default
date: "2026-09-01"
editor_options:
  chunk_output_type: console
  markdown:
    wrap: sentence
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Survival Prediction with SUPPORT2 Dataset

**Dataset URL:** <https://archive.ics.uci.edu/dataset/880/support2>

**Title:** Forecasting 2- and 6-Month Survival of Critically Ill Patients with the Study to Understand Prognoses and Preferences for Outcomes and Risks of Treatment (SUPPORT2) Dataset

**Prediction Problem:** Estimate the probability that a critically ill patient survives 2 and 6 months after hospital admission using routinely collected clinical and demographic data.

**Research question:** How accurately can supervised machine learning models predict 2‑ and 6‑month survival in the SUPPORT2 cohort, and how does their performance compare to the original SUPPORT prognostic model based on discrimination metrics such as AUC and accuracy?

**Hypotheses:**

1.  **Prediction accuracy**\
    The supervised machine learning models will achieve good discrimination (AUC ≈0.80 or higher) for both 2‑ and 6‑month survival and perform at least comparably to the original SUPPORT model on the held‑out test set.
2.  **Model comparison**\
    Gradient boosting and random forest models will show equal or better discrimination than logistic and LASSO regression, reflecting their ability to capture non‑linear and interaction effects in the predictors.
3.  **Prediction horizon**\
    Predictive performance will be slightly higher for 2‑month survival than for 6‑month survival, reflecting the greater uncertainty associated with longer‑term outcomes in critically ill patients.

Formal statistical significance testing (for example, pairwise DeLong tests comparing AUCs across models) was planned but not completed within the available computation time, so results are interpreted descriptively rather than on the basis of p‑values.

## 1. Setup and Package Installations

Install required packages and set seed for reproducibility

```{r}
library(tidyverse)     # For data manipulation
library(dplyr)         # For data manipulation
library(caret)         # For machine learning framework
library(mice)          # For missing data
library(glmnet)        # For regularized regression (Lasso/Ridge)
library(randomForest)  # For random forest models
library(gbm)           # For gradient boosting
library(xgboost)       # For gradient boosting
library(e1071)         # For SVM
library(nnet)          # For neural networks
library(corrplot)      # For correlation plots
library(ggplot2)       # For advanced visualizations and diamonds dataset
library(gridExtra)     # For arranging multiple plots
library(pROC)          # For ROC curves and AUC
library(rpart.plot)    # For decision tree visualization
library(doParallel)    # For efficient ML testing

set.seed(42)

```

## 2. Data loading and Initial Inspection

The SUPPORT2 dataset contains prospectively collected clinical data from 9,105 critically ill adult patients admitted to five United States hospitals between 1989–1991 and 1992–1994.
Patients were enrolled with conditions such as acute respiratory failure, chronic obstructive pulmonary disease, congestive heart failure, liver disease, coma, colon cancer, lung cancer, and multiple organ system failure with or without malignancy or sepsis.
The dataset includes 42 routinely collected variables, covering mortality and survival information, physician survival estimates, demographics (age, sex, race, income, education), hospital cost variables, comorbidities, Do Not Resuscitate (DNR) status, vital signs, laboratory values, and functional outcomes.

```{r}
df <- read.csv("support2.csv")
# View(df) # checked and data were imported correctly

```

For this analysis, the primary outcome variables provided are death at any time during follow‑up (*death*), in‑hospital death (*hospdead*), and functional disability at two months (*sfdm2*), with *slos* (days from study entry to discharge) and *d.time* (days of follow‑up) serving as supporting time‑to‑event measures.
Candidate predictors include demographic variables, baseline disease group and class, comorbidities, cost variables, physiology and severity scores, physician and model‑based survival estimates, DNR information, vital signs, laboratory tests, and activities of daily living measures (*adlp, adls, adlsc*).​

The real‑world goal is to use these routinely collected variables to predict 2‑ and 6‑month survival after hospital admission, thereby informing clinical and end‑of‑life decision‑making for critically ill patients.​

An initial structural review confirms that the dataset contains 9,105 observations and 47 variables, reflecting the 42 predictors plus three primary outcomes and two supporting time variables.
Variable types are a mix of numeric and categorical, missing values are present in several predictors, and no duplicate rows are detected.

```{r}

cat("=== DATA STRUCTURE ===\n")
str(df)  # Shows variable types and first few values

cat("\n=== DATA DIMENSIONS ===\n")
cat("Number of observations:", nrow(df), "\n")
cat("Number of variables:", ncol(df), "\n")

cat("\n=== FIRST FEW ROWS ===\n")
head(df, 10)  # Show first 10 rows

cat("\n=== LAST FEW ROWS ===\n")
tail(df, 10)  # Check last few rows for any issues

# Check for duplicate rows
cat("\n=== DUPLICATE CHECK ===\n")
cat("Number of duplicate rows:", sum(duplicated(df)), "\n")

```

### Initial Cleaning

Because observed 2‑ and 6‑month survival are not directly available, new binary outcome variables are derived from *d.time* and *death*.
For each patient, survival status at 60 and 180 days is coded as 0 for death before the time horizon and 1 for survival to or beyond the horizon; remaining patients are treated as censored and set to missing.

```{r}

df$obs_surv2 <- with(df,
  ifelse(d.time < 60 & death == 1, 0, # died before 60 days
    ifelse(d.time >= 60, 1,           # survived 60 days or died after
      NA                              # alive but <60 days (censored)
    )
  )
)

df$obs_surv2 <- factor(df$obs_surv2,
                            levels = c(0, 1),
                            labels = c("No", "Yes"))

cat("\nObserved 2-month Survival Statistics:\n")
summary(df$obs_surv2)


df$obs_surv6 <- with(df,
  ifelse(d.time < 180 & death == 1, 0, # died before 180 days
    ifelse(d.time >= 180, 1,           # survived 180 days or died after
      NA                              # alive but <180 days (censored)
    )
  )
)

df$obs_surv6 <- factor(df$obs_surv6,
                            levels = c(0, 1),
                            labels = c("No", "Yes"))

cat("\nObserved 6-month Survival Statistics:\n")
summary(df$obs_surv6)

```

To avoid information leakage from post‑baseline outcomes or follow‑up lengths, *death*, *hospdead*, *slos*, and *d.time* are excluded from the predictor set after constructing *obs_surv2* and *obs_surv6*.
The functional disability score *sfdm2* is not used as an outcome because there is no 6‑month analogue and is not retained as a predictor because it reflects post‑admission disability rather than baseline clinical or demographic information.​

Hospital cost variables (*charges, totcst, totmcst*) are removed because they do not align with the clinical research question and may complicate interpretation without adding prognostic value.
The dataset documentation recommends that previously derived prognostic scores (*aps, sps, surv2m, surv6m*) not be used as predictors when developing new models, and advises caution with physician estimates (*prg2m, prg6m*) and DNR variables (*dnr, dnrday*).
In this analysis, *aps* and *sps* are removed, *dnr* and *dnrday* are dropped following guidance that DNR orders should not enter prognostic models, and *prg2m* and *prg6m* are retained for secondary analyses to assess the added value of physician judgment.
The original SUPPORT survival predictions (*surv2m, surv6m*) are preserved only as external benchmarks for model comparison.

The resulting working dataset for EDA and model development includes demographic, diagnostic, physiologic, and functional variables alongside the constructed survival outcomes.

```{r}

#Select relevant columns
support2 <- df[, c("age","sex","dzgroup","dzclass","num.co","edu","income","scoma","avtisst","race","surv2m","surv6m","hday","diabetes","dementia","ca","meanbp","wblc","hrt","resp","temp","pafi","alb","bili","crea","sod","ph","glucose","bun","urine","adlp","adls","adlsc", "prg2m", "prg6m", "obs_surv2","obs_surv6")]

```

## 3. Exploratory Data Analysis

### Summary Statistics

A global summary of the working dataset highlights substantial missingness in several laboratory and functional variables, as well as variation in the distribution and scale of predictors.

```{r}

summary(support2)

```

To quantify missingness, the proportion of missing values is computed for each variable, and a bar plot is used to visualize patterns across predictors.

### Missing Values Analysis

```{r}

cat("\n=== MISSING VALUES ANALYSIS ===\n")
missing_count <- colSums(is.na(support2))
missing_percent <- round(100 * missing_count / nrow(support2), 2)
missing_df <- data.frame(
  Variable = names(missing_count),
  Missing_Count = missing_count,
  Missing_Percent = missing_percent
)
print(missing_df)

# Visualize missing values
if(sum(missing_count) > 0) {
  barplot(missing_percent, 
          main = "Percentage of Missing Values by Variable",
          ylab = "Percentage Missing (%)",
          las = 2,
          col = "coral")
}

```

Variables with less than 10% missingness include *scoma, avtisst, surv2m, surv6m, meanbp, wblc, hrt, resp, temp, crea,* and *sod*.
Moderate missingness (10–30%) is observed for *edu, pafi, prg2m,* and *prg6m*, while several laboratory and ADL measures (*alb, bili, ph, glucose, bun, urine, adlp, adls*) show more than 30% missing values and will require careful handling during preprocessing.

### Target Variable Analysis

The class balance of the constructed survival outcomes is examined to understand the difficulty of the prediction task and to inform later modeling choices.​

```{r}
cat("\n=== TARGET VARIABLE ANALYSIS ===\n")
cat("\nTarget 1: 2-month Survival")

# Frequency table
cat("\nClass Distribution:\n")
table(support2$obs_surv2)
cat("\nClass Proportions:\n")
prop.table(table(support2$obs_surv2))

# Visualize class distribution
par(mfrow = c(1, 2))
barplot(table(support2$obs_surv2), 
        main = "2-month Survival Distribution",
        xlab = "obs_surv2",
        ylab = "Frequency",
        col = c("lightblue", "coral"))

pie(table(support2$obs_surv2), 
    main = "2-month Survival Proportion",
    col = c("lightblue", "coral"))

cat("\nTarget 2: 6-month Survival")

# Frequency table
cat("\nClass Distribution:\n")
table(support2$obs_surv6)
cat("\nClass Proportions:\n")
prop.table(table(support2$obs_surv6))

# Visualize class distribution
par(mfrow = c(1, 2))
barplot(table(support2$obs_surv6), 
        main = "6-month Survival Distribution",
        xlab = "obs_surv6",
        ylab = "Frequency",
        col = c("lightblue", "coral"))

pie(table(support2$obs_surv6), 
    main = "6-month Survival Proportion",
    col = c("lightblue", "coral"))

target_variables <- c("obs_surv2", "obs_surv6") # combined target variable made for later parts of eda
```

As expected in a critically ill cohort, 6‑month survival is lower than 2‑month survival, with a more balanced outcome distribution at 6 months and a modest survival majority at 2 months.

### SUPPORT Model Analysis

To contextualize the new models, the pre‑existing SUPPORT survival predictions at 2 and 6 months (*surv2m, surv6m*) are summarized and visualized.

```{r}

cat("\n=== SUPPORT MODEL ANALYSIS ===\n")
cat("\nTarget 1: SUPPORT 2-month Survival")

# Frequency table
cat("\nDescriptive Statistics:\n")

cat(sprintf("Mean: %.3f\n", mean(support2$surv2m, na.rm = TRUE)))
cat(sprintf("Median: %.3f\n",median(support2$surv2m, na.rm = TRUE)))
cat(sprintf("SD: %.3f\n",sd(support2$surv2m, na.rm = TRUE)))
cat(sprintf("Min: %.3f\n",min(support2$surv2m, na.rm = TRUE)))
cat(sprintf("Max: %.3f\n",max(support2$surv2m, na.rm = TRUE)))

# Visualize class distribution
par(mfrow = c(1, 2))
hist(support2$surv2m, 
        main = "SUPPORT 2-month Survival Distribution",
        xlab = "surv2m",
        ylab = "Frequency",
        col = c("lightblue", "coral"),
        breaks = 20,
        cex.main = 0.8)

boxplot(support2$surv2m, 
        main = "SUPPORT 2-month Survival Proportion",
        col = "lightgreen",
        cex.main = 0.8)

cat("\nTarget 2: SUPPORT 6-month Survival")

# Frequency table
cat("\nDescriptive Statistics::\n")
cat(sprintf("Mean: %.3f\n", mean(support2$surv6m, na.rm = TRUE)))
cat(sprintf("Median: %.3f\n",median(support2$surv6m, na.rm = TRUE)))
cat(sprintf("SD: %.3f\n",sd(support2$surv6m, na.rm = TRUE)))
cat(sprintf("Min: %.3f\n",min(support2$surv6m, na.rm = TRUE)))
cat(sprintf("Max: %.3f\n",max(support2$surv6m, na.rm = TRUE)))

# Visualize class distribution
par(mfrow = c(1, 2))
hist(support2$surv6m, 
        main = "SUPPORT 6-month Survival Distribution",
        xlab = "surv6m",
        ylab = "Frequency",
        col = c("lightblue", "coral"),
        breaks = 20,
        cex.main = 0.8)

boxplot(support2$surv6m, 
        main = "SUPPORT 6-month Survival Proportion",
        col = "lightgreen",
        cex.main = 0.8)

support_models <- c("surv2m", "surv6m") # combined target variable made for later parts of eda
```

Both SUPPORT predictions display a bimodal pattern with a substantial mass near zero, representing patients with very poor prognosis, and a second mode in the higher‑survival range.
At 2 months, predicted survival frequently peaks around 0.8, whereas 6‑month predictions are more dispersed with a peak between 0.6 and 0.8, reflecting increased uncertainty over longer horizons.

### Predictor Variable Analysis

#### Numeric Variable Analysis

Numeric predictors are first identified and then summarized using histograms, boxplots, and descriptive statistics to assess central tendency, spread, skewness, and potential outliers.

```{r}

cat("\n=== NUMERIC VARIABLES ANALYSIS ===\n")
numeric_vars <- setdiff(names(support2)[sapply(support2, is.numeric)], c(target_variables, support_models)) #removes targets and support models from EDA
cat("Numeric variables found:", paste(numeric_vars, collapse = ", "), "\n")

```

Because *diabetes* and *dementia* are binary indicators, they are treated as categorical variables for subsequent analyses and removed from the numeric list.
Several count‑ and scale‑based variables (e.g., *num.co, adlp, adls, adlsc*) represent ordered constructs and may later be converted to ordinal factors, but they are initially explored as numeric.

```{r}

cat_vars <- c("diabetes", "dementia")
numeric_vars <- setdiff(
  names(support2)[sapply(support2, is.numeric)], 
  c(support_models, cat_vars)
  ) #removes categorical vars and support models from EDA
cat("Numeric variables found:", paste(numeric_vars, collapse = ", "), "\n")

```

```{r}
# Distribution plots for numeric variables
# And box plots to identify outliers

par(mar = c(2, 2, 2, 1))
par(mfrow = c(5, 2))
for (var in numeric_vars) {
  hist(support2[[var]], 
       main = paste("Distribution of", var),
       xlab = var,
       col = "lightblue",
       breaks = 30)

  boxplot(support2[[var]], 
          main = var,
          col = "lightgreen")
}

# Detailed statistics for numeric variables
cat("\nDetailed Statistics for Numeric Variables:\n")
for(var in numeric_vars) {
  cat(sprintf("\n--- %s ---\n", var))
  cat(sprintf("Mean: %.2f\n", mean(support2[[var]], na.rm = TRUE)))
  cat(sprintf("Median: %.2f\n", median(support2[[var]], na.rm = TRUE)))
  cat(sprintf("SD: %.2f\n", sd(support2[[var]], na.rm = TRUE)))
  cat(sprintf("Min: %.2f\n", min(support2[[var]], na.rm = TRUE)))
  cat(sprintf("Max: %.2f\n", max(support2[[var]], na.rm = TRUE)))
  cat(sprintf("IQR: %.2f\n", IQR(support2[[var]], na.rm = TRUE)))
  
  # Check for outliers using IQR method
  Q1 <- quantile(support2[[var]], 0.25, na.rm = TRUE)
  Q3 <- quantile(support2[[var]], 0.75, na.rm = TRUE)
  IQR_val <- Q3 - Q1
  outliers <- sum(support2[[var]] < (Q1 - 1.5 * IQR_val) | 
                    support2[[var]] > (Q3 + 1.5 * IQR_val), na.rm = TRUE)
  cat(sprintf("Number of outliers (IQR method): %d (%.2f%%)\n", 
              outliers, 100 * outliers / nrow(support2)))
}

```

Distributions reveal that many physiologic and laboratory variables are right‑skewed, including *num.co, scoma, avtisst, hday, wblc, pafi, alb, bili, crea, glucose, bun, urine,* and the ADL measures, while *meanbp* appears multimodal and *age, resp,* and *temp* show mild skewness.
Most outliers are clinically plausible for a critically ill ICU population and are therefore retained for modeling.

#### Categorical Variable Analysis

Categorical variables are summarized using frequency tables, proportions, and bar plots to characterize the patient population and inform later modeling decisions.

```{r}
cat("\n=== CATEGORICAL VARIABLES ANALYSIS ===\n")
categorical_vars <- setdiff(
  names(support2)[sapply(support2, function(x) is.factor(x) || is.character(x))],
  target_variables
  )
categorical_vars <- c(categorical_vars, cat_vars)
cat("Categorical variables found:", paste(categorical_vars, collapse = ", "), "\n")

# Frequency tables and bar plots
par(mar = c(2, 2, 2, 1))
par(mfrow = c(3, 3))
for(var in categorical_vars) {
  cat(sprintf("\nFrequency table for %s:\n", var))
  print(table(support2[[var]]))
  cat(sprintf("Proportions for %s:\n", var))
  print(prop.table(table(support2[[var]])))
  
  # Bar plot
  barplot(table(support2[[var]]), 
          main = paste("Distribution of", var),
          col = rainbow(length(unique(support2[[var]]))))
}

```

The cohort contains slightly more males than females, is predominantly White, and is dominated by acute renal failure or multiorgan failure with sepsis among disease groups, with smaller representation of cirrhosis and colon cancer.
Most patients do not have diabetes or dementia, and the majority lack a history of cancer, although metastatic disease remains clinically important among those with malignancy.

### Relationship between Predictors and Target

To explore potential predictors of short‑ and medium‑term mortality, numeric variables are compared across survival groups using boxplots, and categorical variables are cross‑tabulated with the 2‑ and 6‑month outcomes.

```{r}
cat("\n=== RELATIONSHIP BETWEEN PREDICTORS AND TARGETS ===\n")

# Both target variables 
# For numeric predictors vs binary target

par(mfrow = c(3, 2), mar = c(4, 4, 2.5, 1))
for(var in numeric_vars) {
  # 2-month survival plot
  boxplot(support2[[var]] ~ support2$obs_surv2,
          main = paste(var, "by 2-month survival"),
          col = c("lightblue", "coral"),
          ylab = var,
          xlab = "obs_surv2",
          cex.main = 0.8)
  
  # 6-month survival plot
  boxplot(support2[[var]] ~ support2$obs_surv6,
          main = paste(var, "by 6-month survival"),
          col = c("lightgreen", "pink"),
          ylab = var,
          xlab = "obs_surv6",
          cex.main = 0.8)
}

# For categorical predictors vs binary target
for(var in categorical_vars[!categorical_vars %in% c("obs_surv2", "obs_surv6")]) {
  
  cat(sprintf("\n=== Variable: %s ===\n", var))
  
  # 2-month survival
  cat("\n--- 2-month survival ---\n")
  tbl2 <- table(support2[[var]], support2$obs_surv2)
  print(tbl2)
  cat("\nProportions (row-wise):\n")
  print(prop.table(tbl2, margin = 1))
  
  # 6-month survival
  cat("\n--- 6-month survival ---\n")
  tbl6 <- table(support2[[var]], support2$obs_surv6)
  print(tbl6)
  cat("\nProportions (row-wise):\n")
  print(prop.table(tbl6, margin = 1))
}

```

Visual inspection suggests that higher coma scores (*scoma*), greater severity (*avtisst*), and certain diagnostic categories (e.g., coma, cancer, multiorgan failure) are associated with lower survival at both time points, whereas age and comorbidity burden show more modest gradients.
ADL measures display an inverse pattern relative to the data dictionary description, with higher values appearing more common in non‑survivors, indicating that the scale is likely coded such that higher scores reflect greater dependence.

#### Note on ADL variables

The data dictionary reports that higher ADL scores indicate better survival, but EDA shows that non‑survivors tend to have higher *adlp, adls,* and *adlsc* values than survivors.
This pattern is consistent with traditional ADL instruments if the coding is reversed, with 0 indicating independence and higher scores indicating increasing dependence, particularly given that surrogate‑reported ADLs capture more severe impairment than patient‑reported scores.
The analysis therefore proceeds under the assumption that lower ADL scores represent greater independence and higher scores greater dependence, attributing the discrepancy to documentation or coding differences.

### Correlation Analysis

Finally, correlations among numeric predictors are examined to identify highly collinear pairs that may require exclusion or feature engineering prior to modeling.

```{r}
if(length(numeric_vars) > 1) {
  cat("\n=== CORRELATION ANALYSIS ===\n")
  
  # Calculate correlation matrix
  cor_matrix <- cor(support2[, numeric_vars], use = "complete.obs")
  print(round(cor_matrix, 3))
  
  # Visualize correlation matrix
  par(mfrow = c(1, 1))
  corrplot(cor_matrix,
         method = "color",
         type = "upper",
         order = "hclust",
         tl.cex = 0.8,
         tl.col = "black")
  
  # Identify highly correlated pairs
  cat("\nHighly correlated variable pairs (|r| > 0.6):\n")
  high_cor <- which(abs(cor_matrix) > 0.6 & abs(cor_matrix) < 1, arr.ind = TRUE)
  if(nrow(high_cor) > 0) {
    for(i in 1:nrow(high_cor)) {
      var1 <- rownames(cor_matrix)[high_cor[i, 1]]
      var2 <- colnames(cor_matrix)[high_cor[i, 2]]
      cor_val <- cor_matrix[high_cor[i, 1], high_cor[i, 2]]
      cat(sprintf("  %s <-> %s: %.3f\n", var1, var2, cor_val))
    }
    cat("\nNote: Consider removing one variable from highly correlated pairs\n")
  } else {
    cat("No highly correlated pairs found.\n")
  }
}

```

Strong correlations are observed between the physician survival estimates (*prg2m, prg6m*) and between the surrogate‑calibrated ADL scores (*adls, adlsc*), leading to retention of *adlsc* and removal of *adls* and later *adlp* due to missingness and redundancy.
Blood urea nitrogen and creatinine are also highly correlated, motivating the later construction of a BUN:creatinine ratio in the preprocessing step to reduce multicollinearity and align with clinical practice.

```{r}

support2 <- support2[, c("age","sex","dzgroup","dzclass","num.co","edu","income","scoma","avtisst","race","surv2m","surv6m","hday","diabetes","dementia","ca","meanbp","wblc","hrt","resp","temp","pafi","alb","bili","crea","sod","ph","glucose","bun","urine","adlsc","prg2m", "prg6m", "obs_surv2","obs_surv6")]

```

### EDA Summary

Overall, the EDA highlights four key issues that shape the modeling strategy: substantial but clinically interpretable missingness, skewed and occasionally multimodal predictor distributions, strong and clinically plausible associations between some predictors and survival, and a small set of highly correlated variables requiring careful selection or transformation.

## 4. Data Preprocessing

Guided by the EDA findings, the dataset was preprocessed to ensure consistent encoding of categorical variables, appropriate handling of missing data, and creation of clinically meaningful derived features before model fitting.

### Data Type Conversion

Categorical predictors stored as character or numeric codes are recast as factors with appropriate levels and labels to support downstream modeling and interpretation.
Income and race fields with blank strings are recoded to missing and then treated as categorical, while binary indicators for diabetes and dementia are explicitly labeled “No” and “Yes.” The number of comorbidities is grouped into clinically interpretable categories, and the surrogate‑calibrated ADL score (*adlsc*) is binned into ordered levels reflecting increasing dependence.

```{r}

support2$sex <- factor(support2$sex)

support2$dzgroup <- factor(support2$dzgroup)
support2$dzclass <- factor(support2$dzclass)

support2$income <- factor(ifelse(trimws(support2$income) == "", NA, support2$income)) #recode "" to NA

support2$race <- factor(ifelse(trimws(support2$race) == "", NA, support2$race)) #recode "" to NA

support2$diabetes <- factor(support2$diabetes,
                            levels = c(0, 1),
                            labels = c("No", "Yes"))

support2$dementia <- factor(support2$dementia,
                            levels = c(0, 1),
                            labels = c("No", "Yes"))

support2$ca <- factor(support2$ca)

support2$num.co_cat <- cut(
    support2$num.co,
    breaks = c(-Inf, 0, 2, 4, Inf),
    labels = c("0", "1-2", "3-4", "5+"),
    right = TRUE
  )
support2$num.co_cat <- factor(support2$num.co_cat, ordered = FALSE)

support2$adlsc_cat <- cut(
  support2$adlsc,
  breaks = c(-0.01, 1, 2, 3, 4, 5, 6, 7, 8),
  labels = c("0","1","2","3","4","5","6","7"),
  right = FALSE,
  include.lowest = TRUE
)
support2$adlsc_cat <- factor(support2$adlsc_cat)

str(support2)

```

### Verify and Encode Any Remaining Categorical Variables

A brief check confirms that all intended categorical variables, including the survival outcomes, are encoded as factors with the expected levels.

```{r}
cat("\n=== VERIFY CATEGORICAL VARIABLE ENCODING ===\n")

cat("Checking if all categorical variables are properly encoded as factors...\n")
for(var in categorical_vars) {
  if(!is.factor(support2[[var]])) {
    support2[[var]] <- as.factor(support2[[var]])
    cat(sprintf("Converted %s to factor\n", var))
  }
}

# Verify target variable is a factor with proper levels
if(!is.factor(support2$obs_surv2)) {
  support2$obs_surv2 <- as.factor(support2$obs_surv2)
  cat(sprintf("Converted target variable '%s' to factor\n", "obs_surv2"))
}

if(!is.factor(support2$obs_surv6)) {
  support2$obs_surv6 <- as.factor(support2$obs_surv6)
  cat(sprintf("Converted target variable '%s' to factor\n", "obs_surv6"))
}

cat(sprintf("\nTarget variable 1 '%s' levels: %s\n", 
            "obs_surv2", 
            paste(levels(support2$obs_surv2), collapse = ", ")))
cat(sprintf("\nTarget variable 2 '%s' levels: %s\n", 
            "obs_surv6", 
            paste(levels(support2$obs_surv6), collapse = ", ")))
cat("All categorical variables are properly encoded!\n")


```

### Handle Missing Values (Simple Drops)

To prevent rare missingness from propagating through imputation, observations with missing values in a small subset of variables (\<0.01% missing) are dropped, and the remaining missingness is addressed in later imputation steps.
The extent of missing data is re‑summarized after this minimal row removal.

```{r}
missing_count <- colSums(is.na(support2))
missing_percent <- round(100 * missing_count / nrow(support2), 2)
missing_df <- data.frame(
  Variable = names(missing_count),
  Missing_Count = missing_count,
  Missing_Percent = missing_percent
)
print(missing_df)
```

```{r}

support2 <- support2 %>% 
  drop_na(scoma, surv2m, surv6m, meanbp, hrt, resp, temp, sod)

missing_count <- colSums(is.na(support2))
missing_percent <- round(100 * missing_count / nrow(support2), 2)
missing_df <- data.frame(
  Variable = names(missing_count),
  Missing_Count = missing_count,
  Missing_Percent = missing_percent
)
print(missing_df)
```

```{r}
support2_clean <- support2
```

### Create Test and Train Set

To ensure fair comparison across models and imputation strategies, a single stratified train–test split is created and then reused for all subsequent analyses.
Stratification is based on the joint distribution of 2‑ and 6‑month survival, preserving the combination of outcomes across training and test sets and facilitating consistent model comparison across time horizons.

```{r}

cat("\n=== SPLITTING DATA INTO TRAIN AND TEST SETS ===\n")
# Create a combined factor for stratification
support2_clean$target_variables <- interaction(support2_clean$obs_surv2, 
                                        support2_clean$obs_surv6, 
                                        drop = TRUE)

# Use stratified sampling to maintain class proportions
set.seed(42)  # For reproducibility
train_index <- createDataPartition(support2_clean$target_variables,
                                   p = 0.75,  # 75% for training
                                   list = FALSE,
                                   times = 1)

train_data <- support2_clean[train_index, ]
test_data  <- support2_clean[-train_index, ]

# Verify class distribution is maintained

for (var in target_variables) {
  cat(sprintf("\nClass distribution for %s in training set:\n", var))
  print(prop.table(table(train_data[[var]])))
  cat(sprintf("\nClass distribution for %s in test set:\n", var))
  print(prop.table(table(test_data[[var]])))
}
```

### Simple Imputation

The first imputation strategy uses a combination of clinically motivated fixed values, grouped medians, and modes to replace missing values.
Fixed values for key laboratory measures follow the dataset documentation and typical clinical norms, while remaining numeric variables are imputed using medians (with sex‑specific medians for *ph* and *glucose*), and *race* and *income* are imputed using their most frequent category.​

```{r}

support2_simple <- support2_clean

#Fixed value imputation
impute_values <- list(
  alb   = 3.5,
  pafi  = 333.3,
  bili  = 1.01,
  crea  = 1.01,
  bun   = 6.51,
  wblc  = 9,
  urine = 2502,
  prg2m = 0.5, #not included in the dataset homepage, 0.5 used for both predictions to 
  prg6m = 0.5  #reduce bias
)

for (var in names(impute_values)) {
  value <- impute_values[[var]]
  
  if (sum(is.na(support2_simple[[var]])) > 0) {
    support2_simple[[var]][is.na(support2_simple[[var]])] <- value
    cat(sprintf("Imputed %s with fixed value: %s\n", var, value))
  }
}

#Grouped median imputation for ph and glucose by sex
group_median_vars <- c("ph", "glucose")

support2_simple <- support2_simple %>%
  group_by(sex) %>%
  mutate(across(all_of(group_median_vars),
                ~ ifelse(is.na(.), median(., na.rm = TRUE), .))) %>%
  ungroup()

#Median imputation for avtsst and edu
median_vars <- c("avtisst", "edu")

for (var in median_vars) {
  if (sum(is.na(support2_simple[[var]])) > 0) {
    med_val <- median(support2_simple[[var]], na.rm = TRUE)
    support2_simple[[var]][is.na(support2_simple[[var]])] <- med_val
    cat(sprintf("Imputed %s with median: %.2f\n", var, med_val))
  }
}


#Mode imputation for race and income
mode_vars <- c("race", "income")

for (var in mode_vars) {
  if (sum(is.na(support2_simple[[var]])) > 0) {
    mode_val <- names(sort(table(support2_simple[[var]]), decreasing = TRUE))[1]
    support2_simple[[var]][is.na(support2_simple[[var]])] <- mode_val
    cat(sprintf("Imputed %s with mode: %s\n", var, mode_val))
  }
}

# Verify no missing values remain
cat("Missing values after imputation:", sum(is.na(support2_simple)), "\n")
```

#### Simple - Feature Engineering

Two derived features are then engineered to capture clinically meaningful patterns and reduce multicollinearity.
A BUN:creatinine ratio summarizes renal function while addressing the strong correlation between *bun* and *crea*, and *pafi* is binned into severity categories based on widely used PaO22/FiO22 cut‑points (<https://litfl.com/pao2-fio2-ratio/>).

```{r}
support2_simple <- support2_simple %>%
  mutate(
    bun_crea_ratio = ifelse(crea > 0, bun / crea, NA_real_)
  )

support2_simple <- support2_simple %>%
  mutate(
    pafi_bin = case_when(
      pafi > 300            ~ "normal",
      pafi >= 200 & pafi <= 300 ~ "mild",
      pafi >= 100 & pafi < 200   ~ "moderate",
      pafi < 100             ~ "severe",
      TRUE                   ~ NA_character_
    )
  )

support2_simple$pafi_bin <- factor(support2_simple$pafi_bin)

```

#### Simple - Train and Test Split

The same train–test indices are then applied to the simply imputed dataset to maintain consistency with the original split.

```{r}

# Split using the same indices
train_data_simple <- support2_simple[train_index, ]
test_data_simple  <- support2_simple[-train_index, ]

```

### Complex Imputation

As a complementary approach, Multiple Imputation by Chained Equations (MICE) is applied to the cleaned dataset to obtain a fully imputed version that better preserves multivariate relationships between predictors.
The imputation model includes both outcomes and a broad set of demographic, clinical, and laboratory predictors, and a single completed dataset (m = 1) is extracted for subsequent modeling given time constraints.

```{r MICE, cache=TRUE}

support2_complex <- support2_clean

# Variables to include in MICE (predictors + outcomes)
cols_for_mice <- c(
  "obs_surv2", "obs_surv6",
  "age", "sex", "dzgroup", "dzclass", "num.co", "edu",
  "income", "scoma", "avtisst", "race", "hday", "diabetes",
  "dementia", "ca", "meanbp", "wblc", "hrt", "resp", "temp",
  "alb", "bili", "sod", "ph", "glucose", "urine",
  "bun", "crea", "pafi", "adlsc",
  "prg2m", "prg6m"
)

cols_for_mice <- intersect(cols_for_mice, names(support2_complex))
mice_data <- support2_complex[, cols_for_mice]

# Ensure outcomes are factors
mice_data$obs_surv2 <- as.factor(mice_data$obs_surv2)
mice_data$obs_surv6 <- as.factor(mice_data$obs_surv6)

# Run MICE once on the full modeling dataset
set.seed(42)
mice_fit <- mice::mice(
  mice_data,
  m = 1,
  maxit = 20
)

mice_complete <- mice::complete(mice_fit, 1)

cat("Total missing after MICE:", sum(is.na(mice_complete)), "\n")

# Put imputed variables back
support2_complex[, cols_for_mice] <- mice_complete
```

#### Complex - Feature Engineering

The same feature engineering steps used in the simple imputation pipeline are applied to the MICE‑imputed dataset, ensuring comparable predictor structures across imputation strategies.
The existing train–test indices are reused to create parallel training and test sets for complex imputation.

```{r}
support2_complex <- support2_complex %>%
  mutate(
    bun_crea_ratio = ifelse(crea > 0, bun / crea, NA_real_)
  )

support2_complex <- support2_complex %>%
  mutate(
    pafi_bin = case_when(
      pafi > 300            ~ "normal",
      pafi >= 200 & pafi <= 300 ~ "mild",
      pafi >= 100 & pafi < 200   ~ "moderate",
      pafi < 100             ~ "severe",
      TRUE                   ~ NA_character_
    )
  )

support2_complex$pafi_bin <- factor(support2_complex$pafi_bin)

```

#### Complex - Train and Test Split

```{r}

# Split using the same indices
train_data_complex <- support2_complex[train_index, ]
test_data_complex  <- support2_complex[-train_index, ]

```

#### Imputation lists

To simplify downstream modeling, the imputed datasets are stored in a structured list separating simple and complex strategies and their corresponding train and test subsets.

```{r}

imputations <- list(
  simple = list(train = train_data_simple,  test = test_data_simple),
  complex = list(train = train_data_complex, test = test_data_complex)
)
```

### Define Predictor Variables

Three predictor sets are defined to support different modeling scenarios and to assess the incremental value of physician survival estimates.
The base set includes only baseline demographic, diagnostic, physiologic, and functional measures, while extended sets add *prg2m* for 2‑month models and *prg6m* for 6‑month models, respectively.

```{r}

predictor_vars <- c( "age", "sex", "dzgroup", "dzclass", "num.co_cat", "edu",
                     "income", "scoma", "avtisst", "race", "hday", "diabetes",
                     "dementia", "ca", "meanbp", "wblc", "hrt", "resp", "temp",
                     "alb", "bili", "sod", "ph", "glucose", "urine",
                     "bun_crea_ratio", "pafi_bin", "adlsc_cat"
                     )

predictor_vars_prg2m <- c( "age", "sex", "dzgroup", "dzclass", "num.co_cat", "edu",
                     "income", "scoma", "avtisst", "race", "hday", "diabetes",
                     "dementia", "ca", "meanbp", "wblc", "hrt", "resp", "temp",
                     "alb", "bili", "sod", "ph", "glucose", "urine",
                     "bun_crea_ratio", "pafi_bin", "adlsc_cat", "prg2m"
                     )

predictor_vars_prg6m <- c( "age", "sex", "dzgroup", "dzclass", "num.co_cat", "edu",
                     "income", "scoma", "avtisst", "race", "hday", "diabetes",
                     "dementia", "ca", "meanbp", "wblc", "hrt", "resp", "temp",
                     "alb", "bili", "sod", "ph", "glucose", "urine",
                     "bun_crea_ratio", "pafi_bin", "adlsc_cat", "prg6m"
                     )

predictor_sets <- list(
  base = predictor_vars,
  prg2m = predictor_vars_prg2m,
  prg6m = predictor_vars_prg6m
)
```

### Note on Outliers and Scaling

Outliers in key predictors were reviewed during EDA and found to be clinically plausible for a critically ill ICU population, so no additional trimming or winsorization is applied.
Because the primary models (logistic regression, LASSO, random forest, gradient boosting) are either scale‑invariant or internally standardized, external feature scaling is not performed.

## 5. Model Building

4 models will be built:

-   **Logistic regression**: transparent baseline, easy interpretation.

-   **Lasso logistic regression**: handles correlated predictors, does variable selection.

-   **Random forest**: strong nonlinear, interaction‑capable benchmark.

-   **Gradient boosting**: often best performance in structured tabular data.

These four supervised learning models are developed and compared using a common train–test split and repeated cross‑validation: logistic regression, LASSO logistic regression, random forest, and gradient boosting.
These models collectively provide a transparent baseline, penalized regression with embedded variable selection, and two non‑linear tree‑based approaches capable of capturing complex interactions in tabular clinical data.

To satisfy the requirement of evaluating multiple models and at least two cross‑validated approaches while remaining computationally feasible, imputation strategies are compared first using logistic regression and the better‑performing approach is then carried forward to LASSO, random forest, and gradient boosting.
The extended predictor sets including physician survival estimates (*prg2m, prg6m*) are only applied to the best‑performing model to assess the incremental contribution of physician judgment without incurring prohibitive run times.

### Model Prep

#### Cross Validation

All models use the same repeated 10‑fold cross‑validation scheme, with ROC as the primary optimization metric, to obtain robust internal performance estimates and ensure fair comparison across algorithms.

```{r}

ctrl <- trainControl(
  method = "repeatedcv",      # Cross-validation with repeats (slight improvement)
  number = 10,                # 10 folds
  repeats = 5,                # 5 repeats
  summaryFunction = twoClassSummary,  # For binary classification metrics
  classProbs = TRUE,          # Need class probabilities for ROC
  savePredictions = "final",  # Save predictions for analysis
  verboseIter = FALSE,         # Set to TRUE to see progress
  allowParallel = FALSE  
)
```

#### Model Results

A unified results table is initialized to capture key evaluation metrics for each model, imputation strategy, and outcome (2‑ and 6‑month survival), including cross‑validated training ROC, test accuracy, sensitivity, specificity, AUC, and training time.

The roc_list object is used to store ROC curves for each model–imputation–outcome combination.
This structure was originally introduced to enable formal pairwise AUC comparison using tests such as DeLong’s method (for example, comparing each model to the SUPPORT benchmark or to the best‑performing model).
If additional computation time were available, these stored ROC objects could be passed to functions like roc.test() to quantify whether observed differences in AUC between models are statistically meaningful, but in this report results are interpreted descriptively without inferential testing.

```{r}

# Initialize results storage
# Create an empty data frame to store performance metrics for different models
model_results <- data.frame(
  Imputation = character(),         # "simple" or "complex"
  Target = character(),              # obssurv2 or obssurv6
  Model = character(),      # Model name (e.g., Logistic Regression, Random Forest)
  Accuracy_Train = numeric(),        # Accuracy on the training dataset
  Accuracy_Test = numeric(),         # Accuracy on the test dataset
  Sensitivity_Test = numeric(),      # True positive rate on the test set
  Specificity_Test = numeric(),      # True negative rate on the test set
  AUC_Test = numeric(),              # Area Under the ROC Curve on the test set
  Training_Time = numeric(),         # Time taken to train the model (in seconds)
  stringsAsFactors = FALSE           # Keep text columns as character, not factors
)

# Store ROC objects for later significance tests
roc_list <- list()
```

### Model 1: Logistic Regression

Logistic regression serves as an interpretable baseline model and is fitted under both imputation strategies and for both survival horizons using the base predictor set.
Test‑set performance across imputation methods is then compared to select the imputation strategy used in subsequent models.

```{r m1, cache=TRUE}
for (imp_name in names(imputations)) {
  for (target_var in target_variables) {
    cat("\n\n============================\n")
    cat("LOGISTIC REGRESSION -", 
        "Imputation:", imp_name, 
        "| Predictor set: predictor_vars",
        "| Target:", target_var, "\n")
    cat("============================\n")
    
    train_data <- imputations[[imp_name]]$train
    test_data  <- imputations[[imp_name]]$test
    
    formula <- as.formula(
      paste(target_var, "~", paste(predictor_vars, collapse = " + "))
    )
    
    start_time <- Sys.time()
    set.seed(42)
    
    log_model <- train(
      formula,
      data   = train_data,
      method = "glm",
      family = "binomial",
      trControl = ctrl,
      metric    = "ROC"
    )
    
    end_time  <- Sys.time()
    log_time  <- as.numeric(difftime(end_time, start_time, units = "secs"))
    
    log_pred_test      <- predict(log_model, test_data)
    log_pred_test_prob <- predict(log_model, test_data, type = "prob")
    
    log_cm_test <- confusionMatrix(
      log_pred_test,
      test_data[[target_var]],
      positive = "Yes"
    )
    
    log_roc_test <- roc(
      test_data[[target_var]],
      log_pred_test_prob[, "Yes"],
      levels = c("No", "Yes")
    )
    log_auc_test <- as.numeric(auc(log_roc_test))
    
    model_name <- "M1: LogReg"
    ps_name    <- "predictor_vars"
    roc_key    <- paste0(model_name, "_", imp_name, "_", ps_name, "_", target_var)
    roc_list[[roc_key]] <- list(
      roc          = log_roc_test,
      Model        = model_name,
      Imputation   = imp_name,
      PredictorSet = ps_name,
      Target       = target_var
    )
    
    model_results <- rbind(
      model_results,
      data.frame(
        Imputation       = imp_name,
        PredictorSet     = ps_name,
        Target           = target_var,
        Model            = model_name,
        Accuracy_Train   = max(log_model$results$ROC),
        Accuracy_Test    = as.numeric(log_cm_test$overall["Accuracy"]),
        Sensitivity_Test = as.numeric(log_cm_test$byClass["Sensitivity"]),
        Specificity_Test = as.numeric(log_cm_test$byClass["Specificity"]),
        AUC_Test         = log_auc_test,
        Training_Time    = log_time,
        stringsAsFactors = FALSE
      )
    )
  }
}

```

#### Best imputation method:

For both 2‑ and 6‑month survival, logistic regression with simple imputation yields slightly higher AUC and accuracy than with MICE‑based complex imputation, so the simple imputation pipeline is selected for all subsequent models.​

### Model 2: Lasso Logistic Regression

LASSO logistic regression extends the baseline model by applying an l1 penalty to perform variable selection and shrinkage, helping to manage multicollinearity and high‑dimensional predictor sets.
This model is trained on the simply imputed dataset using the base predictor set for both outcomes, with internal centering and scaling.

```{r m2, cache=TRUE}
cl <- makePSOCKcluster(parallel::detectCores() - 1)
registerDoParallel(cl)

# Use simple imputation and predictor_vars only
imp_name   <- "simple"
ps_name    <- "predictor_vars"
train_data <- imputations[[imp_name]]$train
test_data  <- imputations[[imp_name]]$test

for (target_var in target_variables) {
  cat("\n\n============================\n")
  cat("LASSO LOGISTIC REGRESSION -", 
      "Imputation:", imp_name, 
      "| Predictor set:", ps_name,
      "| Target:", target_var, "\n")
  cat("============================\n")
  
  # build formula for this target
  formula <- as.formula(
    paste(target_var, "~", paste(predictor_vars, collapse = " + "))
  )
  
  # fit model
  start_time <- Sys.time()
  set.seed(42)
  
  lasso_model <- train(
    formula,
    data = train_data,
    method = "glmnet",
    family = "binomial",
    trControl = ctrl,          # consider ctrl$allowParallel <- FALSE if glmnet is fussy
    tuneLength = 30,
    metric = "ROC",
    preProcess = c("center", "scale")
  )
  
  end_time <- Sys.time()
  lasso_time <- as.numeric(difftime(end_time, start_time, units = "secs"))
  
  cat("Training complete!\n")
  print(lasso_model)
  
  # Variable importance
  cat("\nVariable Importance (Lasso):\n")
  lasso_importance <- varImp(lasso_model)
  print(lasso_importance)
  
  # Predictions
  lasso_pred_test      <- predict(lasso_model, test_data)
  lasso_pred_test_prob <- predict(lasso_model, test_data, type = "prob")
  
  # Evaluate
  lasso_cm_test <- confusionMatrix(
    lasso_pred_test,
    test_data[[target_var]],
    positive = "Yes"
  )
  cat("\nTest Set Performance:\n")
  print(lasso_cm_test)
  
  lasso_roc_test <- roc(
    test_data[[target_var]],
    lasso_pred_test_prob[, "Yes"],
    levels = c("No", "Yes")
  )
  lasso_auc_test <- as.numeric(auc(lasso_roc_test))
  cat(sprintf("\nTest Set AUC: %.4f\n", lasso_auc_test))
  
  # Store ROC
  model_name <- "M2: Lasso"
  roc_key <- paste0(model_name, "_", imp_name, "_", ps_name, "_", target_var)
  roc_list[[roc_key]] <- list(
    roc          = lasso_roc_test,
    Model        = model_name,
    Imputation   = imp_name,
    PredictorSet = ps_name,
    Target       = target_var
  )
  
  # Store results
  model_results <- rbind(
    model_results,
    data.frame(
      Imputation       = imp_name,
      PredictorSet     = ps_name,
      Target           = target_var,
      Model            = model_name,
      Accuracy_Train   = max(lasso_model$results$ROC),
      Accuracy_Test    = as.numeric(lasso_cm_test$overall["Accuracy"]),
      Sensitivity_Test = as.numeric(lasso_cm_test$byClass["Sensitivity"]),
      Specificity_Test = as.numeric(lasso_cm_test$byClass["Specificity"]),
      AUC_Test         = lasso_auc_test,
      Training_Time    = lasso_time,
      stringsAsFactors = FALSE
    )
  )
}

stopCluster(cl)
registerDoSEQ()

```

### Model 3: Random Forest

Random forest provides a flexible non‑linear benchmark that can capture complex interactions and handles mixed‑type predictors without extensive preprocessing.
Using the simple imputation pipeline and base predictor set, a tuned random forest is fitted for each outcome, with a small grid over *mtry* and fixed values for the number of trees and tree size constraints.

```{r m3, cache=TRUE}
cl <- makePSOCKcluster(parallel::detectCores() - 1)
registerDoParallel(cl)

# Use simple imputation and predictor_vars only
imp_name   <- "simple"
ps_name    <- "predictor_vars"
train_data <- imputations[[imp_name]]$train
test_data  <- imputations[[imp_name]]$test

for (target_var in target_variables) {
  cat("\n\n============================\n")
  cat("RANDOM FOREST -", 
      "Imputation:", imp_name, 
      "| Predictor set:", ps_name,
      "| Target:", target_var, "\n")
  cat("============================\n")
  
  # Build formula with predictor_vars
  formula <- as.formula(
    paste(target_var, "~", paste(predictor_vars, collapse = " + "))
  )
  
  n_features <- length(predictor_vars)
  
  # mtry grid (smart tuning)
  rf_grid <- expand.grid(
    mtry = unique(pmax(2, round(c(sqrt(n_features), n_features/3, n_features/2))))
  )
  
  # Train model
  start_time <- Sys.time()
  set.seed(42)
  
  rf_model <- train(
    formula,
    data = train_data,
    method = "rf",
    trControl = ctrl,
    tuneGrid = rf_grid,
    ntree = 300,
    maxnodes = 30,
    nodesize = 5,
    importance = TRUE,
    metric = "ROC"
  )
  
  end_time <- Sys.time()
  rf_time <- as.numeric(difftime(end_time, start_time, units = "secs"))
  
  cat("Training complete!\n")
  print(rf_model)
  
  # Variable importance
  cat("\nVariable Importance (Random Forest):\n")
  rf_importance <- varImp(rf_model)
  print(rf_importance)
  
  # Predictions
  rf_pred_test      <- predict(rf_model, test_data)
  rf_pred_test_prob <- predict(rf_model, test_data, type = "prob")
  
  # Evaluation
  rf_cm_test <- confusionMatrix(
    rf_pred_test,
    test_data[[target_var]],
    positive = "Yes"
  )
  rf_roc_test <- roc(
    test_data[[target_var]],
    rf_pred_test_prob[, "Yes"],
    levels = c("No","Yes")
  )
  rf_auc_test <- as.numeric(auc(rf_roc_test))
  
  cat("\nTest Set Performance:\n")
  print(rf_cm_test)
  cat(sprintf("\nTest Set AUC: %.4f\n", rf_auc_test))
  
  # Store ROC
  model_name <- "M3: RF"
  roc_key <- paste0(model_name, "_", imp_name, "_", ps_name, "_", target_var)
  roc_list[[roc_key]] <- list(
    roc          = rf_roc_test,
    Model        = model_name,
    Imputation   = imp_name,
    PredictorSet = ps_name,
    Target       = target_var
  )
  
  # Store results
  model_results <- rbind(
    model_results,
    data.frame(
      Imputation       = imp_name,
      PredictorSet     = ps_name,
      Target           = target_var,
      Model            = model_name,
      Accuracy_Train   = max(rf_model$results$ROC),
      Accuracy_Test    = as.numeric(rf_cm_test$overall["Accuracy"]),
      Sensitivity_Test = as.numeric(rf_cm_test$byClass["Sensitivity"]),
      Specificity_Test = as.numeric(rf_cm_test$byClass["Specificity"]),
      AUC_Test         = rf_auc_test,
      Training_Time    = rf_time,
      stringsAsFactors = FALSE
    )
  )
}

stopCluster(cl)
registerDoSEQ()

```

### Model 4: Gradient Boosting

Gradient boosting (GBM) is included as a high‑capacity ensemble model that often delivers strong performance in structured tabular data.
Using the simple imputation and base predictor set, GBM models are tuned over a grid of tree counts, interaction depths, and learning rates for each outcome.

```{r m4, cache=TRUE}
cl <- makePSOCKcluster(parallel::detectCores() - 1)
registerDoParallel(cl)

# Use simple imputation and predictor_vars only
imp_name   <- "simple"
ps_name    <- "predictor_vars"
train_data <- imputations[[imp_name]]$train
test_data  <- imputations[[imp_name]]$test

for (target_var in target_variables) {
  cat("\n\n============================\n")
  cat("GRADIENT BOOSTING -", 
      "Imputation:", imp_name, 
      "| Predictor set:", ps_name, 
      "| Target:", target_var, "\n")
  cat("============================\n")
  
  predictor_variables <- predictor_vars
  
  formula <- as.formula(
    paste(target_var, "~", paste(predictor_variables, collapse = " + "))
  )
  
  gbm_grid <- expand.grid(
    n.trees = seq(100, 1000, by = 100),
    interaction.depth = c(1, 3, 5),
    shrinkage = c(0.01, 0.05, 0.1),
    n.minobsinnode = 10
  )
  
  start_time <- Sys.time()
  set.seed(42)
  
  gbm_model <- train(
    formula,
    data = train_data,
    method = "gbm",
    trControl = ctrl,
    tuneGrid = gbm_grid,
    metric = "ROC",
    verbose = FALSE
  )
  
  end_time <- Sys.time()
  gbm_time <- as.numeric(difftime(end_time, start_time, units = "secs"))
  
  cat("Training complete!\n")
  print(gbm_model)
  
  # Variable importance
  gbm_importance <- varImp(gbm_model)
  print(gbm_importance)
  
  # Predictions
  gbm_pred_test_prob <- predict(gbm_model, test_data, type = "prob")
  gbm_pred_test      <- predict(gbm_model, test_data)
  
  # Evaluation
  gbm_cm_test <- confusionMatrix(
    gbm_pred_test,
    test_data[[target_var]],
    positive = "Yes"
  )
  gbm_roc_test <- roc(
    test_data[[target_var]],
    gbm_pred_test_prob[, "Yes"],
    levels = c("No","Yes")
  )
  gbm_auc_test <- as.numeric(auc(gbm_roc_test))
  
  cat("\nTest Set Performance:\n")
  print(gbm_cm_test)
  cat(sprintf("\nTest Set AUC: %.4f\n", gbm_auc_test))
  
  # Store ROC
  model_name <- "M4: GBM"
  roc_key <- paste0(model_name, "_", imp_name, "_", ps_name, "_", target_var)
  roc_list[[roc_key]] <- list(
    roc          = gbm_roc_test,
    Model        = model_name,
    Imputation   = imp_name,
    PredictorSet = ps_name,
    Target       = target_var
  )
  
  # Store results (use GBM metrics, not logistic ones)
  model_results <- rbind(
    model_results,
    data.frame(
      Imputation       = imp_name,
      PredictorSet     = ps_name,
      Target           = target_var,
      Model            = model_name,
      Accuracy_Train   = max(gbm_model$results$ROC),
      Accuracy_Test    = as.numeric(gbm_cm_test$overall["Accuracy"]),
      Sensitivity_Test = as.numeric(gbm_cm_test$byClass["Sensitivity"]),
      Specificity_Test = as.numeric(gbm_cm_test$byClass["Specificity"]),
      AUC_Test         = gbm_auc_test,
      Training_Time    = gbm_time,
      stringsAsFactors = FALSE
    )
  )
}

# Stop parallel
stopCluster(cl)
registerDoSEQ()

```

#### Best Model

Because gradient boosting emerged as the best‑performing model based on test‑set AUC and accuracy at both time horizons, an additional set of GBM models was fitted to evaluate the incremental value of physician survival estimates.
Specifically, for 2‑month survival, a GBM model including the base predictors plus prg2m was trained, and for 6‑month survival a corresponding GBM model including prg6m was estimated, using the same simple imputation pipeline, train–test split, and cross‑validation settings as the primary GBM models.

To limit runtime while still assessing the added value of physician estimates, two streamlined gradient boosting models were fit as add‑on analyses.
For 2‑month survival, a GBM model was retrained using the simple imputation pipeline, the same train–test split and cross‑validation framework, and an extended predictor set that included *prg2m* in addition to the original clinical and demographic variables.
An analogous GBM model was then fit for 6‑month survival using *prg6m* alongside the base predictors, but with a reduced tuning grid and simplified 10‑fold cross‑validation (without repeats) to keep computational costs manageable.

```{r GBM+prg2m, cache=TRUE}

cat("\n\n=== ADD-ON: GBM WITH prg2m FOR 2-MONTH SURVIVAL ===\n")

imp_name   <- "simple"
ps_name    <- "prg2m"
train_data <- imputations[[imp_name]]$train
test_data  <- imputations[[imp_name]]$test

target_var <- "obs_surv2"

formula_gbm_prg2m <- as.formula(
  paste(target_var, "~", paste(predictor_sets[[ps_name]], collapse = " + "))
)

# Lighter tuning grid to reduce runtime
gbm_grid_prg2m <- expand.grid(
  n.trees = c(100, 300),      # fewer trees
  interaction.depth = 3,      # single moderate depth
  shrinkage = 0.05,           # single learning rate
  n.minobsinnode = 10
)

# Simpler CV: 10-fold, no repeats (do not change global ctrl)
ctrl_light <- trainControl(
  method = "cv",
  number = 10,
  summaryFunction = twoClassSummary,
  classProbs = TRUE,
  savePredictions = "final",
  verboseIter = FALSE,
  allowParallel = FALSE
)

start_time <- Sys.time()
set.seed(42)

gbm_prg2m_model <- train(
  formula_gbm_prg2m,
  data = train_data,
  method = "gbm",
  trControl = ctrl_light,
  tuneGrid = gbm_grid_prg2m,
  metric = "ROC",
  verbose = FALSE
)

end_time <- Sys.time()
gbm_prg2m_time <- as.numeric(difftime(end_time, start_time, units = "secs"))

gbm_prg2m_pred_prob <- predict(gbm_prg2m_model, test_data, type = "prob")
gbm_prg2m_pred      <- predict(gbm_prg2m_model, test_data)

gbm_prg2m_cm <- confusionMatrix(
  gbm_prg2m_pred,
  test_data[[target_var]],
  positive = "Yes"
)

gbm_prg2m_roc <- roc(
  test_data[[target_var]],
  gbm_prg2m_pred_prob[, "Yes"],
  levels = c("No","Yes")
)
gbm_prg2m_auc <- as.numeric(auc(gbm_prg2m_roc))

cat("\nTest Set Performance (GBM + prg2m, 2-month):\n")
print(gbm_prg2m_cm)
cat(sprintf("Test AUC: %.4f\n", gbm_prg2m_auc))

# Store in roc_list and model_results so comparison code still works
model_name <- "M4: GBM + prg2m"
roc_key <- paste0(model_name, "_", imp_name, "_", ps_name, "_", target_var)
roc_list[[roc_key]] <- list(
  roc          = gbm_prg2m_roc,
  Model        = model_name,
  Imputation   = imp_name,
  PredictorSet = ps_name,
  Target       = target_var
)

model_results <- rbind(
  model_results,
  data.frame(
    Imputation       = imp_name,
    PredictorSet     = ps_name,
    Target           = target_var,
    Model            = model_name,
    Accuracy_Train   = max(gbm_prg2m_model$results$ROC),
    Accuracy_Test    = as.numeric(gbm_prg2m_cm$overall["Accuracy"]),
    Sensitivity_Test = as.numeric(gbm_prg2m_cm$byClass["Sensitivity"]),
    Specificity_Test = as.numeric(gbm_prg2m_cm$byClass["Specificity"]),
    AUC_Test         = gbm_prg2m_auc,
    Training_Time    = gbm_prg2m_time,
    stringsAsFactors = FALSE
  )
)

```

```{r GBMprg6m, cache=TRUE}

cat("\n\n=== ADD-ON: GBM WITH prg6m FOR 6-MONTH SURVIVAL ===\n")

imp_name   <- "simple"
ps_name    <- "prg6m"
train_data <- imputations[[imp_name]]$train
test_data  <- imputations[[imp_name]]$test

target_var <- "obs_surv6"

formula_gbm_prg6m <- as.formula(
  paste(target_var, "~", paste(predictor_sets[[ps_name]], collapse = " + "))
)

# Same lightweight grid
gbm_grid_prg6m <- expand.grid(
  n.trees = c(100, 300),
  interaction.depth = 3,
  shrinkage = 0.05,
  n.minobsinnode = 10
)

# Reuse the same light CV setup
ctrl_light <- trainControl(
  method = "cv",
  number = 10,
  summaryFunction = twoClassSummary,
  classProbs = TRUE,
  savePredictions = "final",
  verboseIter = FALSE,
  allowParallel = FALSE
)

start_time <- Sys.time()
set.seed(42)

gbm_prg6m_model <- train(
  formula_gbm_prg6m,
  data = train_data,
  method = "gbm",
  trControl = ctrl_light,
  tuneGrid = gbm_grid_prg6m,
  metric = "ROC",
  verbose = FALSE
)

end_time <- Sys.time()
gbm_prg6m_time <- as.numeric(difftime(end_time, start_time, units = "secs"))

gbm_prg6m_pred_prob <- predict(gbm_prg6m_model, test_data, type = "prob")
gbm_prg6m_pred      <- predict(gbm_prg6m_model, test_data)

gbm_prg6m_cm <- confusionMatrix(
  gbm_prg6m_pred,
  test_data[[target_var]],
  positive = "Yes"
)

gbm_prg6m_roc <- roc(
  test_data[[target_var]],
  gbm_prg6m_pred_prob[, "Yes"],
  levels = c("No","Yes")
)
gbm_prg6m_auc <- as.numeric(auc(gbm_prg6m_roc))

cat("\nTest Set Performance (GBM + prg6m, 6-month):\n")
print(gbm_prg6m_cm)
cat(sprintf("Test AUC: %.4f\n", gbm_prg6m_auc))

model_name <- "M4: GBM + prg6m"
roc_key <- paste0(model_name, "_", imp_name, "_", ps_name, "_", target_var)
roc_list[[roc_key]] <- list(
  roc          = gbm_prg6m_roc,
  Model        = model_name,
  Imputation   = imp_name,
  PredictorSet = ps_name,
  Target       = target_var
)

model_results <- rbind(
  model_results,
  data.frame(
    Imputation       = imp_name,
    PredictorSet     = ps_name,
    Target           = target_var,
    Model            = model_name,
    Accuracy_Train   = max(gbm_prg6m_model$results$ROC),
    Accuracy_Test    = as.numeric(gbm_prg6m_cm$overall["Accuracy"]),
    Sensitivity_Test = as.numeric(gbm_prg6m_cm$byClass["Sensitivity"]),
    Specificity_Test = as.numeric(gbm_prg6m_cm$byClass["Specificity"]),
    AUC_Test         = gbm_prg6m_auc,
    Training_Time    = gbm_prg6m_time,
    stringsAsFactors = FALSE
  )
)

```

### SUPPORT Models

The original SUPPORT prognostic model is included as an external benchmark by evaluating its 2‑ and 6‑month survival predictions (*surv2m, surv6m*) on the same held‑out test set used for the machine learning models.
For each time horizon, a ROC curve and AUC are computed, and the resulting AUC values are added to the consolidated results table to allow direct descriptive comparison with the newly trained models.
No formal statistical tests are applied to compare AUCs; differences are interpreted descriptively due to computational constraints.

```{r support models}

ref_test <- imputations[["simple"]]$test

## 2-month SUPPORT
roc_support_2m <- roc(
  response  = ref_test$obs_surv2,
  predictor = ref_test$surv2m,
  levels    = c("No","Yes")
)
auc_support_2m <- as.numeric(auc(roc_support_2m))

## 6-month SUPPORT
roc_support_6m <- roc(
  response  = ref_test$obs_surv6,
  predictor = ref_test$surv6m,
  levels    = c("No","Yes")
)
auc_support_6m <- as.numeric(auc(roc_support_6m))

# Store SUPPORT ROC objects in roc_list for plotting
roc_list[["SUPPORT_surv2m_obssurv2"]] <- list(
  roc          = roc_support_2m,
  Model        = "SUPPORT (surv2m)",
  Imputation   = NA,
  PredictorSet = NA,
  Target       = "obs_surv2"
)

roc_list[["SUPPORT_surv6m_obssurv6"]] <- list(
  roc          = roc_support_6m,
  Model        = "SUPPORT (surv6m)",
  Imputation   = NA,
  PredictorSet = NA,
  Target       = "obs_surv6"
)

# Append SUPPORT benchmarks to the results table
model_results <- rbind(
  model_results,
  data.frame(
    Imputation       = NA,
    PredictorSet     = NA,
    Target           = "obs_surv2",
    Model            = "SUPPORT (surv2m)",
    Accuracy_Train   = NA,
    Accuracy_Test    = NA,
    Sensitivity_Test = NA,
    Specificity_Test = NA,
    AUC_Test         = auc_support_2m,
    Training_Time    = NA,
    stringsAsFactors = FALSE
  ),
  data.frame(
    Imputation       = NA,
    PredictorSet     = NA,
    Target           = "obs_surv6",
    Model            = "SUPPORT (surv6m)",
    Accuracy_Train   = NA,
    Accuracy_Test    = NA,
    Sensitivity_Test = NA,
    Specificity_Test = NA,
    AUC_Test         = auc_support_6m,
    Training_Time    = NA,
    stringsAsFactors = FALSE
  )
)

```

## 6. Model Comparison and Selection

The fitted models are compared using the consolidated results table, focusing on test‑set accuracy, AUC, sensitivity, specificity, and training time for each outcome and model.
Visual summaries are then used to highlight relative performance across algorithms and to support selection of a single primary model for interpretation and benchmarking against the SUPPORT predictions.

```{r}

cat("\n\n########## MODEL COMPARISON ##########\n\n")

#Display results table

cat("=== MODEL PERFORMANCE COMPARISON ===\n") 
print(model_results)

#Sort by test accuracy

model_results_sorted <- model_results[order(-model_results$Accuracy_Test), ] 
cat("\nModels ranked by Test Accuracy:\n") 
print(model_results_sorted)

#Visualize model comparison

#par(mfrow = c(2, 2))

#1. Test Accuracy Comparison
par(mar = c(10, 4, 4, 2))  
barplot(model_results$Accuracy_Test,
        names.arg = model_results$Model, las = 2, col = rainbow(nrow(model_results)), main = "Test Set Accuracy by Model", ylab = "Accuracy", ylim = c(0, 1))

#2. AUC Comparison
par(mar = c(10, 4, 4, 2))  
barplot(as.numeric(model_results$AUC_Test),
        names.arg = model_results$Model, las = 2, col = rainbow(nrow(model_results)), main = "Test Set AUC by Model", ylab = "AUC", ylim = c(0, 1))

#3. Training Time Comparison
par(mar = c(10, 4, 4, 2))  
barplot(model_results$Training_Time,
        names.arg = model_results$Model, las = 2, col = rainbow(nrow(model_results)), main = "Training Time by Model", ylab = "Time (seconds)")

#4. Sensitivity Comparison
par(mar = c(10, 4, 4, 2))  
barplot(model_results$Sensitivity_Test,
        names.arg = model_results$Model, las = 2,
        col = rainbow(nrow(model_results)),
        main = "Test Set Sensitivity by Model",
        ylab = "Sensitivity",
        ylim = c(0, 1))


```

```{r}
# 5. ROC Curves Comparison

cat("\n=== ROC CURVE PANELS BY OUTCOME ===\n")

# Helper to plot ROC panel for a given target
plot_roc_panel <- function(target_name, support_key, main_title) {
  # Subset results for this target (exclude rows without AUC)
  df_t <- subset(model_results, Target == target_name & !is.na(AUC_Test))
  
  # Exclude SUPPORT rows when ranking ML models
  df_ml <- subset(df_t, !grepl("^SUPPORT", Model))
  
  if (nrow(df_ml) == 0) {
    warning(sprintf("No ML models found for target '%s'; skipping panel.", target_name))
    return(invisible(NULL))
  }
  
  # Select top 3 models by AUC
  top_ml <- head(df_ml[order(-df_ml$AUC_Test), ], 3)
  
  # Get SUPPORT ROC object
  if (!support_key %in% names(roc_list)) {
    stop(sprintf("Support ROC key '%s' not found in roc_list.", support_key))
  }
  roc_support <- roc_list[[support_key]]$roc
  
  # Open plot and draw SUPPORT curve
  plot(
    roc_support,
    col = "black",
    lwd = 2,
    main = main_title,
    legacy.axes = TRUE
  )
  
  # Colors for up to 3 top ML models
  ml_cols <- c("blue", "red", "green")
  
  # Add ROC for each top ML model
  for (i in seq_len(nrow(top_ml))) {
    m  <- top_ml$Model[i]
    im <- top_ml$Imputation[i]
    ps <- top_ml$PredictorSet[i]
    
    key <- paste0(m, "_", im, "_", ps, "_", target_name)
    if (!key %in% names(roc_list)) {
      warning(sprintf("ROC object for key '%s' not found in roc_list; skipping.", key))
      next
    }
    
    roc_obj <- roc_list[[key]]$roc
    plot(roc_obj, col = ml_cols[i], lwd = 2, add = TRUE)
  }
  
  # Build legend labels
  support_idx <- which(grepl("^SUPPORT", df_t$Model))
  if (length(support_idx) == 0) {
    warning(sprintf("No SUPPORT row found for target '%s' in model_results.", target_name))
    support_label <- "SUPPORT"
  } else {
    support_label <- sprintf("SUPPORT (AUC = %.2f)", df_t$AUC_Test[support_idx[1]])
  }
  
  legend_labels <- c(support_label)
  legend_cols   <- c("black")
  
  for (i in seq_len(nrow(top_ml))) {
    legend_labels <- c(
      legend_labels,
      sprintf("%s (AUC = %.2f)", top_ml$Model[i], top_ml$AUC_Test[i])
    )
    legend_cols <- c(legend_cols, ml_cols[i])
  }
  
  legend(
    "bottomright",
    legend = legend_labels,
    col    = legend_cols,
    lwd    = 2,
    cex    = 0.7
  )
}

# Two-panel layout: 2-month and 6-month survival
par(mfrow = c(1, 2))

# 2-month survival panel
plot_roc_panel(
  target_name = "obs_surv2",
  support_key = "SUPPORT_surv2m_obssurv2",
  main_title  = "ROC Curves – 2-month Survival"
)

# 6-month survival panel
plot_roc_panel(
  target_name = "obs_surv6",
  support_key = "SUPPORT_surv6m_obssurv6",
  main_title  = "ROC Curves – 6-month Survival"
)

```

```{r}
#Select best model based on AUC

best_model_idx <- which.max(model_results$AUC_Test)
best_model_name <- model_results$Model[best_model_idx] 
cat(sprintf("\n=== BEST MODEL: %s ===\n", best_model_name)) 
cat(sprintf("Test Accuracy: %.4f\n", model_results$Accuracy_Test[best_model_idx]))
cat(sprintf("Test Sensitivity: %.4f\n", model_results$Sensitivity_Test[best_model_idx])) 
cat(sprintf("Test Specificity: %.4f\n", model_results$Specificity_Test[best_model_idx]))
cat(sprintf("Test AUC: %.4f\n", model_results$AUC_Test[best_model_idx]))
```

### Discussion

Gradient boosting with physician survival estimates provided the strongest overall discrimination and accuracy at both 2‑ and 6‑month horizons, outperforming logistic regression, LASSO, and random forest as well as the original SUPPORT scores.
Incorporating *prg2m* and *prg6m* into the GBM models yielded meaningful gains in AUC, test accuracy, and sensitivity compared with GBM models using clinical predictors alone, supporting the value of physician judgement in short‑term mortality prediction.​

#### Overall model performance

Across all models, performance was generally higher for 2‑month than 6‑month survival, with modest trade‑offs between sensitivity and specificity.
For 2‑month survival, GBM with base predictors achieved test accuracy of 0.77 and AUC of 0.81, while logistic and LASSO regression were close behind with slightly lower AUCs but similar accuracies; random forest performed slightly worse on both metrics.
For 6‑month survival, GBM again led the base‑predictor models (AUC ≈ 0.81), with logistic and LASSO clustered just below and random forest lagging somewhat in both accuracy and discrimination.​

Compared with the SUPPORT prognostic model, all four machine‑learning models trained on the SUPPORT2 predictors achieved similar or better AUCs for both time horizons.
SUPPORT’s AUCs of about 0.79 for 2 months and 0.78 for 6 months were matched or exceeded by logistic and LASSO regression and clearly surpassed by GBM, indicating that modest gains in discrimination are achievable with contemporary supervised learning while retaining the same underlying predictor information.​

#### Effect of physician estimates

Adding physician 2‑month survival estimates (*prg2m*) to the GBM model for 2‑month survival increased AUC from 0.81 to 0.83 and accuracy from 0.77 to 0.79, with a corresponding improvement in sensitivity and only a small change in specificity.
Similarly, including *prg6m* for 6‑month survival raised GBM AUC from 0.81 to 0.82 and accuracy from 0.73 to 0.75, again with higher sensitivity and acceptable specificity, suggesting that physician judgement contributes independent prognostic information beyond routinely collected clinical variables.​

These add‑on GBM models achieved these gains with substantially lower training time than the fully tuned base GBM models, owing to the lighter tuning grid and simpler 10‑fold cross‑validation used for the physician‑augmented analyses.
The ROC panels confirm visually that the GBM + physician models dominate the other curves across most of the false‑positive rate range for both 2‑ and 6‑month survival, reinforcing their role as the best‑performing models in this exercise.

### Conclusion

The first hypothesis was supported, with GBM and penalized regression achieving higher AUCs than SUPPORT for both short‑ and medium‑term survival.

The second hypothesis was only partially supported: GBM consistently outperformed logistic and LASSO regression, but random forest did not offer clear advantages in this dataset.​

The third hypothesis was supported for the gradient boosting model at both time horizons.
GBM models augmented with *prg2m* and *prg6m* delivered the highest AUC and accuracy of all tested models, indicating that physician judgement contains prognostically useful information not fully captured by baseline clinical variables and strengthening the case for integrating structured clinician estimates into risk‑prediction tools in this setting.\
